---
title: "DATA101 Week3"
author: "Abdirisak Mohamed"
date: "Spring 2018"
output:
  html_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    keep_tex: yes
    toc: yes
    toc_depth: 2
geometry: right=2.5in
---

```{r cache=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=5.5, fig.width=6, collapse=TRUE, comment="", prompt=TRUE)
options(width=63)
```


# Functions

From the book **R Programming for Data Science** 
   
- Writing functions is a core activity of an R programmer. It represents the key step of the        transition from a mere “user” to a developer who creates new functionality for R. 

- Functions are often used to encapsulate a sequence of expressions that need to be executed         numerous times, perhaps under slightly different conditions. 

- Functions are also often written when code must be shared with others or the public.
   

## Functions in R

-Functions are defined using the function() directive and are stored as R objects

-Functions can be passed as arguments to other functions. This is very handy for the various
apply funtions, like lapply() and sapply().

-Functions can be nested, so that you can define a function inside of another function

## Examples
```{r}
increment <- function(x) {
  x<-x +1
  x
}

increment(x=5)

increment(5)

new_increment <- increment
new_increment(x=5)
```

Another example
```{r}
increment_1 <- function(x) {
  y <- x + 1
  return(list(x_original=x, x_increment=y))
}

increment_1(x=5)

z <- increment_1(x=5)

```


Default function arguments
```{r}
expon_1 <- function(x, y, exponent) {
  if(exponent == 2) {
                      x^y }
              else {
      
                    y^x }
}
  
expon_1(x=2, y=3, exponent = 2 )

expon_1(x=2, y=3, exponent = 1 )

# Now we make the argument exponent default with exponent = 2

expon_new <- function(x, y, exponent = 2) {
  if(exponent == 2) {
                      x^y }
              else {
      
                    y^x }
}
  
expon_new(x=2, y=3)

expon_new(x=2, y=3, exponent = 1 )

```

## The ... Argument

- The ... argument indicates a variable number of arguments that are usually passed on to other    functions

```{r}
myplot <- function(x, y, type = "l", ...) {
plot(x, y, type = type, ...) ## Pass '...' to 'plot' function
}
```

## After the ... Argument

- Any arguments that appear after ... on the argument list must be named
  explicitly and cannot be partially matched (e.g, se instead of sep ) or matched positionally if   the default values are not going to be used.
  
- Fro example, paste() doesn't know in advance how many character vectors
  will be passed to the function by the user. So the first argument to the function is ....


```{r}
args(paste)

paste("a", "b", sep = ":")

paste("a", "b", se = ":") # partial sep 
```

## Coding Standards for R

- Always use text files / text editor: RStudio instead of Microsoft Word

- Indent your code

- Limit the width of your code

- Limit the length of individual functions

- Comment your code

## Binary Formats for Storing Data

- RData file is a binary file for saving R objects

Create 'a' and 'b' 
```{r}
a <- data.frame(x = rnorm(100), y = runif(100))
b <- c(3, 4.4, 1 / 3)
```

Save 'a' and 'b' to a file and remove from your environment (workspace)
```{r}
save(a, b, file = "mydata.rda")

rm(a,b) # rm(list=ls()) for removing everything

```

Load 'a' and 'b' into your environment
```{r}

load("mydata.rda")
b

```



# Subsetting R Objects

- There are three operators that can be used to extract subsets of R objects.

    -The [ operator always returns an object of the same class as the original. It can be used to
      select multiple elements of an object
      
    -The [[ operator is used to extract elements of a list or a data frame. It can only be used        to extract a single element and the class of the returned object will not necessarily be a       list or data frame.
    
    -The $ operator is used to extract elements of a list or data frame by literal name. Its           semantics are similar to that of [[.


## Subsetting Vectors

```{r}
x <- c("name", "height", "weight")

x[1]           #  first element
x[2]           #  second element

x[1:2]         # first and second

x[c(1, 3)]  # first and third
x[-c(1, 3)] #  all elements EXCEPT first and third
```

Using logical vectors for subsetting

```{r}
x > "name"

x[x > "name"]

y<-x[x != "name"]

y

```

## Subsettng Matrices

```{r}
x <- matrix(1:6, 2, 3)
x

x[1,2]
x[2,1]
x[1, ]  ## Extract the first row 
x[ ,2]  ## Extract the second column 

```



```{r}
colnames(x) <- c("age", "height", "weight")

x
```

```{r}
x[ , c("height", "weight")]

x[c(FALSE, TRUE), c("height", "weight")]

x[2, c("height", "weight")]


```

## Subsettng Matrices

```{r}
odd <- (x %% 2) == 1

odd

x[odd]

x[c(3, 5, 1)]
```


## Subsetting Lists

```{r}

x <- list(measures = 1:3, cities=c("Rockville", "Gaithersburg", "Germantown"), 
          statement=c(FALSE, TRUE, NA))

x[[1]]
x[["measures"]]
x$measures

```

```{r}
x[[c(3,1)]] # Get the first element of the third element
x[[3]][1]   # Get the first element of the third element
```

```{r}
x[c(3,1)] #Get the third and the first elements
```

## Subsetting Data Frames

```{r}
x <- data.frame(measures = 1:3, cities=c("Rockville", "Gaithersburg", "Germantown"), 
          statement=c(FALSE, TRUE, NA))

x[[1]]
x[["measures"]]
x$measures
```

```{r}
x[[c(3,1)]]
x[[3]][1]
```

```{r}
x[c(3,1)]
```



```{r}
x <- data.frame(measures = 1:3, cities=c("Rockville", "Gaithersburg", "Germantown"), 
          statement=c(FALSE, TRUE, NA))

x[1, ]
x[ ,3]
x[ ,"cities"]
```

```{r}
x[1:2, ]
x[ ,2:3]
```


Remark: It happens many times that R converts character strings to factors.
See the previous example

You can prevent it as follows:

```{r}
x <- data.frame(measures = 1:3, cities=c("Rockville", "Gaithersburg", "Germantown"), 
          statement=c(FALSE, TRUE, NA), stringsAsFactors=FALSE)

x[ ,"cities"]
class(x[ ,"cities"])
```


## Removing NA Values (Missing Values)

A common task in data analysis is removing missing values (NAs).

```{r}
x <- c(1, 2, NA, 4, NA, 5) 
bad <- is.na(x) 
bad 
```

```{r}
 x[!bad] 
```

What if there are multiple R objects and you want to take the subset with no missing values in any of those objects?

```{r}
x <- c(1, 2, NA, 4, NA, 5) 
y <- c("a", "b", NA, "d", NA, "f") 
good <- complete.cases(x, y) 
good 
```

```{r}
x[good] 
y[good] 
```


```{r}
data("airquality")  # data("airquality", package="datasets")
head(airquality)
dim(airquality)
str(airquality)
```

```{r}
 good <- complete.cases(airquality) 
head(airquality[good,])
```

```{r}
which(is.na(airquality$Ozone))
sum(is.na(airquality$Ozone))
```

## Subsetting by Matching

```{r}
letters
vowels <- c("a", "e", "i", "o", "u")

letters %in% vowels
which(letters %in% vowels)

letters[which(letters %in% vowels)]
```


See the *R Programming for Data Science* , chapter:  "Subsetting R Objects" and
also [*Advanced R*](http://adv-r.had.co.nz/Subsetting.html) 



# Managing DataFrames with the dplyr package

## What is Tidy Data

From [*Tidy Data,  Wickham H (2014)*](https://www.jstatsoft.org/article/view/v059i10)

- Tidy datasets are easy to manipulate, model and visualize, and have a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.

- A dataset is a collection of values, usually either numbers (if quantitative) or strings (if qualitative). Values are organized in two ways. Every value belongs to a variable and an observation. A variable contains all values that measure the same underlying attribute (like height, temperature, duration) across units. An observation contains all values measured on the same unit (like a person, or a day, or a race) across attributes.


## Reshape

See [*Reshape,  Wickham H (2007)*](https://www.jstatsoft.org/article/view/v021i12)

Tidy data can be shaped in wide and long formats.

- Wide format data have one observation in each row and a seperate column for each variable.
- Wide data uses a column for each variable.

- Long format data will have data in multiple rows,
- Long format data have a column for variables names and another column for the values of the     variables.

 We take a look at functions of the `reshape2` package.

```{r}

#library("datasets")
#data(airquality, package="datasets")
data(airquality)
names(airquality)
dim(airquality)
```

```{r}
library("reshape2")
library("dplyr")
airquality <- tbl_df(airquality)
class(airquality)
```

Let us check some of the records

```{r}
head(airquality)
```

```{r}
tail(airquality)
```

 Now we use the `melt` function to change into long format

```{r}
long_data <- melt(airquality)
head(long_data)
```

```{r}
tail(long_data)
```
we did not specify Id-variable. Let use Month and Day as Id-variables.

ID variables identify individual rows of data.
 

```{r}
long_data <- melt(airquality, id.vars = c("Month", "Day"))
head(long_data)
```

```{r}
tail(long_data)
```

Now we use `dcast` to change from long to wide format


```{r}
wide_data <- dcast(long_data, Month + Day ~ variable)
head(wide_data)
```

```{r}
tail(wide_data)
```

## The dplyr Package

- The dplyr package was developed by Hadley Wickham of RStudio and is an optimized and distilled version of his plyr package.

- Everything dplyr does could already be done with base R, but it greatly simplifies existing functionality in R. 

## dplyr Grammar

- Some of the key “verbs” provided by the dplyr package are

• select: return a subset of the columns of a data frame, using a flexible notation 

• filter: extract a subset of rows from a data frame based on logical conditions 

• arrange: reorder rows of a data frame 

• rename: rename variables in a data frame 

• mutate: add new variables/columns or transform existing variables 

• summarise / summarize: generate summary statistics of different variables in the data frame, possibly within strata 

• %>%: the “pipe” operator is used to connect multiple verb actions together into a pipeline

## All dplyr functions have the following properties:

1. The first argument is a data frame. 

2. The subsequent arguments describe what to do with the data frame specified in the first argument,and you can refer to columns in the data frame directly without using the $operator (just use the column names). 

3. The return result of a function is a new data frame 

4. Data frames must be tidy

```{r}
 install.packages("dplyr", repos = "http://cran.us.r-project.org")
 library(dplyr)
```

[chicago Pollution](http://www.biostat.jhsph.edu/~rpeng/leanpub/rprog/chicago_data.zip)

The dataset is airpollution and temperature data for the city of Chicago in the U.S
```{r}
chicago<-readRDS("chicago.rds")
 dim(chicago) 
 str(chicago)
```
 
## Select

The select() function can be used to select columns of a data frame that you might need. 

```{r}
subset_1<-select(chicago, city:dptp) 
head(subset_1)
```

Omit columns
```{r}
subset_2<-select(chicago, -(city:dptp))
head(subset_2)
```

Variable names based on patterns
```{r}
subset_3 <- select(chicago, ends_with("2")) 
head(subset_3)

subset_4<- select(chicago, starts_with("d")) 
str(subset_4)
```

## Filter

The filter() function is used to extract subsets of rows from a data frame

```{r}
#  extract the rows of the chicago data frame where the levels of PM2.5 are greater than 30
chic.f <- filter(chicago, pm25tmean2 > 30) 
str(chic.f)

summary(chic.f$pm25tmean2)
```

Extract the rows where PM2.5 is greater than 30 and temperature is greater than 80 degrees Fahrenheit.
```{r}
chic.f <- filter(chicago, pm25tmean2 > 30 & tmpd > 80) 
select(chic.f, date, tmpd, pm25tmean2) 
```


## Arrange

The arrange() function is used to reorder rows of a data frame according to one of the variables/columns. Reordering rows of a data frame (while preserving corresponding order of other columns) is normally a pain to do in R.

```{r}
   chicago <- arrange(chicago, date)
```

The first and last rows
```{r}
  head(select(chicago, date, pm25tmean2), 3) 
  tail(select(chicago, date, pm25tmean2), 3) 
```

Descending order
````{r}
  chicago <- arrange(chicago, desc(date))
   head(select(chicago, date, pm25tmean2), 3) 
   tail(select(chicago, date, pm25tmean2), 3) 
```

## Rename

Let us use the function rename() to change the names of variables.
The column dptp stands for the dew point temperature and the pm25tmean2 column is for the PM2.5 data
```{r}
 head(chicago[, 1:5], 3) 
  chicago <- rename(chicago, dewpoint = dptp, pm25 = pm25tmean2) 
   head(chicago[, 1:5], 3) 
```

## Mutate

The mutate() function computes transformations of variables in a data frame
Let us create pm25detrend variable that subtracts the mean from the pm25 variable.

```{r}
chicago <- mutate(chicago, pm25detrend = pm25 - mean(pm25, na.rm = TRUE)) 
head(chicago) 

# There is also transmute() which drops all non-transformed variables. See the book.
```

## Group_by

-The group_by() function is used to generate summary statistics from the data frame within strata defined by a variable

-Suppose we want the average annual level of PM2.5. Then the stratum is the year and can be obtained from the date variable. 

- The function summarize() is used with the group_by() function. 

- We use the function as.POSIXlt() for date-time conversion and the function mutate() to create a variable for `year`

```{r}
  chicago <- mutate(chicago, year = as.POSIXlt(date)$year + 1900)

# A new data frame that data frame that groups the original data frame by year

 years <- group_by(chicago, year)
 

```

-Summary statistics for each year 
```{r}
 summarize(years, pm25 = mean(pm25, na.rm = TRUE),  o3 = max(o3tmean2, na.rm = TRUE), no2 = median(no2tmean2, na.rm = TRUE)) 
```

## %>%

The pipeline operater %>% strings operations from left to right
```{r}
mutate(chicago, year = as.POSIXlt(date)$year + 1900)  %>% group_by(year) %>% summarize(pm25 = mean(pm25, na.rm = TRUE),  o3 = max(o3tmean2, na.rm = TRUE), no2 = median(no2tmean2, na.rm = TRUE)) 
```
