---
title: "Introduction to R"
author: "Abdirisak Mohamed"
date: "Spring 2018"
output:
  html_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    keep_tex: yes
    toc: yes
    toc_depth: 2
geometry: right=2.5in
---

```{r cache=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=5.5, fig.width=6, collapse=TRUE, comment="", prompt=TRUE)
options(width=63)
```


# Basics of R

RStudio is an integrated development environment for R. One of the advantages of R is that you can start off as a user. It has an interactive console where you can enter commands. It is similar to a calculator.

## Arithmetic
```{r}
4 + 5
```

```{r}
 1 + 2*3 - 4^2
```

```{r}
 1 + (2*3) - (4^2)
```

## R Objects

R has five atomic (basic) classes of objects:

* Character
* Numeric (real numbers)
* Integer
* Complex
* Logic (TRUE/FALSE)

## Examples (the <- is the assignment operator)
```{r}
name<-"Rockville"
x<-sqrt(2)
n<-8L
z<-3 + 4i
statement<-FALSE
```

The # character signals the start of  a comment.  # and everything after it  are not evaluated

```{r}
y<-9 # not evaluated
```

We used the function sqrt() above. It is one of many built-in functions in R.

[Built-In Functions](http://www.statmethods.net/management/functions.html)

## Print. You can either auto-print or use the function print()
```{r}
y
print(y)
```

## Integer sequence
```{r}
z<-1:10
z
x <- seq(from=0, to=10, by=2)
x
length(x)
```

## Vectors: The most basic object

A vector can only contain elements of the same class
```{r}
w<-c("name", "age", "gender")
x<-c("TRUE", "FALSE")
y<-c(1.5, 2.7)
z<-1:10

```

The function vector() creates a vector
```{r}
x<-vector("numeric", length = 4)
x
```

Index of a vector starts at 1 not 0
```{r}
x<-c(7,8,9)
x[1]
x[0] # numeric vector of length zero (i.e., empty)
```

## Implicit Coersion:
If you put elements of different classes in a vector, R coerces them into one class
```{r}
x <- c(2.9, "b")   ## character 
class(x)
y <- c(5, FALSE)    ## numeric 
class(y)
z <- c("b", TRUE)  ## character
class(z)

```

## Explicit Coersion
You can be explicitly coerce a vector from one class to another using exisitng as.* functions
```{r}
x <- 1:3  
x
class(x) 
as.numeric(x) 
as.logical(x)  # 0 = FALSE, other integers = TRUE
as.character(x)

# You get NAs as the conversion is not supported
y <- c("a", "b", "c") 
y
as.numeric(y) 
as.logical(y) 

# Logical to numeric
z<-c(TRUE, FALSE, TRUE)
z
as.numeric(z)
as.character(z)
```

## Matrices
Matrices are vectors with  dimensions. The dimension attribute is a vector v = c(nrow, ncol) of integers. Like vecors, matrices contain elements of the same class.
```{r}
my_mat <- matrix(nrow = 3, ncol = 4) 
my_mat    
dim(my_mat)  
attributes(my_mat) 
```

Matrices are filled column-wise. You can change it by setting byrow = TRUE.
```{r}
mat_2<- matrix(1:6, nrow = 2, ncol = 3) 
mat_2

# Fill by rows
mat_3<- matrix(1:6, nrow = 2, ncol = 3, byrow = TRUE) 
mat_3
```

Adding a dimension attribute to a vector makes it a matrix
```{r}
x<-1:6
x
dim(x)<-c(2,3)
x
```

# Column-binding and Row-binding 
```{r}
x<-1:6
x
y<-11:16
y
c_mat<-cbind(x,y)
c_mat
r_mat<-rbind(x,y)
r_mat
```

## Lists
Lists are like vectors but they can contain elements of different classes
```{r}
x <- list(1:6, c("name","gender", "age"), FALSE) 
x
```

Accessing elements of a list
```{r}
x[[3]]
x[[2]][3]
x[[1]][5]
```

# Factors
Factors represent categorical variables
```{r}
x <- factor(c("January", "April", "January", "June", "April", "April"))
x
table(x)
unclass(x)
# You can set the order of the levels
y <- factor(c("January", "April", "January", "June", "April", "April"), 
            levels = c("January", "April", "June"))
y
```

## Missing Values

NA stands for Not Available and  NaN stands for Not a Number.

The function is.na() is used to test if an element or object is NA.

Similarly, is.nan() is  for testing NaN.

NA values have a class,e.g.,  integer NA, character NA. 

Every NaN value is also NA value but the converse does not hold

```{r}
 x<-c(3, 5, NA, "a", TRUE) 
is.na(x) 
is.nan(x) 
y<-c(3, 5, NA, "a", NaN) 
is.na(y) 
is.nan(y) 

```

##Data Frames 
Data Frames contain data in tabular form.

Data Frames are a special type of list; every member  of the list has  the same length.

The columns are the members of the list.

Data frames can (in contrast to matrices) contain different classes of objects

Data frames have the attribute  row.names

If you want to read a file as a data frame, use read.table() or read.csv().

Data Frames can be converted to a matrix by the function data.matrix() 

```{r}
df <- data.frame( age = c(20,30,25, 23), 
                  class = c("Freshman", "Senior" , "Junior", "Sophomore") )
df
nrow(df) 
ncol(df) 
```

## Names

Vectors can have names
```{r}
vec<-c(25, 5.5, 150)
vec
names(vec)
names(vec)<-c("age", "height", "weight")
vec
names(vec)

```

Lists with names
```{r}
z<-list(age = c(20,30,25, 23), class = c("Freshman", "Senior" , "Junior", "Sophomore"))
z
```

Accessing a list by names
```{r}
z$age
z$class
```

Matrices with names
```{r}
my_mat <- matrix(1:6, nrow = 2, ncol = 3) 
dimnames(my_mat) <- list(c("a", "b"), c("c", "d", "e")) 
my_mat
```

Accessing names
```{r}
names(df)
colnames(df)
rownames(df)
row.names(df)
```